% Introduce the problem to be tackled
% Use this as an opportunity to talk about the inspiration of the project
% Contextualise the problem being tackled by the project
% Explain why the problem is worthy of study
% Outline the structure of the document and provide a framework for the reader of the dissertation. 

\chapter{Project Introduction} % (fold)
\label{cha:project_introduction}
This section will:
\begin{itemize}
    \item \textbf{Introduce the Project:} A brief overview of the inspiration for the project, and how that materialised as the project itself.
    \item \textbf{Overview of Initial Concept}: The initial concept for the system.
    \item \textbf{The Final System:} How that initial concept came to materialise as the final system.
    \item \textbf{Document Overview:} An overview of what the dissertation aims to achieve, and it structure. 
\end{itemize}

This project aims to create a framework and toolchain for the formal verification of embedded, executable Domain-Specific Languages.
The hope is to allow the generation of provably correct, tailored programming languages that can encapsulate domain logic inside another program.
These languages will be compiled through the use of a Haskell metacompiler (a compiler that generates a compiler) that takes both syntax and semantics for the Domain-Specific Language, generating executable libraries for embedding the DSL.

\section{An Introduction to Domain-Specific Languages} % (fold)
\label{sec:an_introduction_to_domain_specific_languages}
Modern software systems are increasing in complexity, with everything from flight computers to business software growing \citep{dvorak2009nasa}.
As these systems grow, the domain logic is becoming integrated throughout the systems, leading to increased levels of coupling throughout the code.
This means that there is an increased risk that changes to the software create modes of incorrect operation \citep{khawar2001developing}.\\

In reaction to the dispersion of domain logic throughout modern software systems, Domain-Specific Languages (DSLs) have seen ``a significant uptick in interest'' \citep{fowler2010domain}. 
A DSL is a limited, application-specific language that is integrated with another software system, providing a custom encoding of domain rules and logic that can often go beyond the syntactic constraints of the host language \citep{Mernik:2005:DDL:1118890.1118892}.
In the use of a DSL, software systems are able to centralise their domain logic using a common syntax, making it simple to change rules for the operation of the program.\\

The term `Domain-Specific Language' is a broad one, encompassing a variety of embedded syntaxes for defining domain logic.
Fowler defines `internal' and `external' DSLs.
An internal DSL is written in the host language and exposed via an API, also known as a fluent interface.
\cite{fowler2010domain} defines an external DSL, in contrast, as a language parsed independently of the host language.\\

In addition to their structure, Domain-Specific Languages can also be defined in terms of their `executability' \citep{Mernik:2005:DDL:1118890.1118892}.
Many DSLs are programming languages with `well-defined execution semantics'.
Examples of this are \LaTeX, a macro language for \TeX, and Template Haskell, which allows manipulation of Haskell code itself using Haskell \citep{Sheard:2002:TMH:581690.581691}.
As a counterpoint, some DSLs only define configuration information for a software system, having no executable semantics of their own.

% section an_introduction_to_domain_specific_languages (end)

\section{The Need for Correct DSLs} % (fold)
\label{sec:the_need_for_correct_dsls}
Initial inspiration for this project was engendered by the observation of a financial technology company using the general-purpose programming language OCaml for specifying logic at runtime.
As a Turing-complete programming language, this offered vastly greater expressive power than necessary for the domain in which it was applied.
This meant an increase in the potential for bugs in the domain logic, as suggested by \cite{subramanyam2003empirical}.
If a capable DSL could be designed that could have formally verified properties, then it would be a much better fit to the domain while reducing the capacity for bugs.\\

While the scope of common DSLs is necessarily very broad, encompassing everything from executable program logic to statically defined configuration, this project aims to focus on a small subset of DSLs.
As formal correctness only applies to executable programs, this project will focus on external, executable DSLs. 
While analysis could benefit internal DSLs as well, they are already covered by traditional code review mechanisms, and would be hard to integrate into this project.\\

Even with the ``detailed analysis and structuring of the application domain'' \citep{van2000domain} performed when initially defining the DSL, it can be difficult to verify the correctness of the resulting language. 
While current methods can utilise code-review to attempt to analyse the correctness of the DSL implementation, a lack of understanding of the domain by the application developers may cause logic errors to be missed. 
This is because the domain logic is often embodied by expert knowledge, as found by \cite{studer1998knowledge}. 
It can hence be difficult to verify the language implementation with those domain experts, through code-review, due to their lack of understanding of the implementing language.
As core domain logic, often integral to the program, is encapsulated by the DSL, it is instrumental that it be formally correct.\\

This project aims to provide a solution for the sixth step of DSL design as proposed by \cite{van2000domain}: ``design and implement a compiler that translates DSL programs''. 
While the logic that the DSL aims to encode can be verified with domain experts, it is often difficult to verify the resulting implementation.
As the language semantics are integral to the operation of the DSL, providing the semantic analysis and verification as part of the toolchain will allow the automatic generation of DSLs.
The generation from specification greatly reduces the potential for errors introduced during the implementation.
An automated system offers benefits over manual review, as the domain experts can work with the DSL designers to encode the correct semantics before it becomes an executable program.

% section the_need_for_correct_dsls (end)

\section{Verification of DSL Semantics} % (fold)
\label{sec:verification_of_dsl_semantics}
While it would be desirable to design a system for verification of arbitrary DSL semantics, there are certain classes of problem in semantic verification which are not decidable \citep{abdulla1994undecidable}. \\

This is most elegantly expressed in the duality of data and codata, or recursion and corecursion. 
While recursively defined structures are decomposed from a state into certain `base-cases', providing a well-defined termination property, corecursion aims to start from a `base-case' and build data from there \citep{hinze2010reasoning}.
As corecursion is dual to recursion, codata is the potentially infinite dual to data, which is necessarily finite. \\

If the program semantics are limited to those that can be defined recursively rather than corecursively; that is, via structural induction rather than coinduction, then termination can be proven for such programs \citep{nordstrom1988terminating}.\\

In the limited domain of recursively defined program semantics, this project aims to build on the work of \cite{doh2001composing}, drawing on the \textit{PlanCompS} project for specifying modular programming language semantics.
It will also draw upon the work of Godel's `System-T', a theory of arithmetic for finite types that allows reasoning about arbitrarily-sized, finite data structures via structural induction \cite{girard1989proofs,alves2010godel}.\\

The aim is to define a set of what Mosses terms `funcons' (\textit{fundamental constructs}); a set of semantic building blocks that can all be proven to terminate \cite{Churchill:2014:RCS:2577080.2577099}.
Alongside a metalanguage that is able to specify both syntax and semantics for the DSL, the project aims to create a toolchain for verifying the DSL semantics and producing an executable library for the DSL.\\

Using the metalanguage definition of the syntax and semantics for the DSL, the project toolchain will act as a metacompiler \cite{Mandell:1966:MDA:800267.810785}.
This means it will produce an executable for the DSL via metacompilation and compilation steps.
The provided semantics will allow decomposition, through structural induction, into the funcons, allowing the termination properties to be proved.
The funcons and associated program code can then be composed to generate the program that matches the semantics defined for the DSL.\\

% section verification_of_dsl_semantics (end)

\section{The Novel Contribution of this Project} % (fold)
\label{sec:the_novel_contribution_of_this_project}
This project aims to make a novel research contribution through the creation of two main components:
\begin{itemize}
    \item \textbf{A DSL Metalanguage:} A metalanguage for the specification of the syntax of a DSL, along with its associated semantics. 
    The syntactic notation will likely be based upon the Extended Backus-Naur Form, a metasyntax notation for expressing languages whose syntax matches a context-free grammar. 
    The corresponding semantic notations will likely be based on the conventional big-step operational semantics \citep{Schmidt:2003:PLS:1074100.1074733} devised for automatic compiler generation \citep{diehl1996semantics}.
    \item \textbf{A Metacompiler:} This metacompiler program will take a DSL description in the above metalanguage, and a program in the described DSL, and produce a working executable from this. 
    This may operate in a single-step or multipass process, depending on what makes for the most simple metacompiler architecture.
\end{itemize}

Through limiting scope to just recursively defined semantic constructs, and omitting consideration of corecursive constructs, this project will be able to produce a useful metacompiler for a set of language semantics within which arbitrary DSLs can be constructed. 

% section the_novel_contribution_of_this_project (end)

\section{Project Objectives} % (fold)
\label{sec:project_objectives}
The main objectives of this project are as follows:
\begin{itemize}
    \item Conduct research on methods for the syntactic description of fundamental program syntax and semantics.
    \item Use the knowledge gained from this research to develop the \textit{Absol}\footnote{Automatic Builder for Semantically Organised Languages: The name of the metacompiler toolchain} metalanguage to describe the syntax and semantics of an arbitrary domain-specific language.
    \item Define a set of basic \textit{funcons} as fundamental building blocks for programs with purely recursively defined semantics. 
    These building blocks will have trivially true termination properties.
    \item Describe methods for semantic decomposition, allowing the program semantics specified using \textit{Absol} to be decomposed to the \textit{funcons}, allowing termination proof for the semantics via structural induction (they are all total functions, and compositions of total functions are also total).
    Structurally recursive semantic definitions allow us to decompose the semantics to the \textit{funcons}, as long as we can express all DSL semantics in terms of these building blocks. 
    \item Development of a basic DSL using the metalanguage.
    \item Development of a minimum working example for the metacompiler using hand-implemented semantics based on the \textit{Absol} metalanguage.
    \item Development of a working metacompiler that can ingest a description of the DSL's syntax and semantics using \textit{Absol}, as well as a program in the DSL itself, and produce a working executable program.
    \item Creation of a more-capable and useful DSL with additional capabilities.
    \item Evaluate the metacompiler in terms of real-world utility and theoretical contribution based on the requirements and the theoretical and practical impact in relation to the literature surveyed.
\end{itemize}


% section project_objectives (end)

% chapter project_introduction (end)
